An item response theory analysis of the Matrix Reasoning Item Bank (MaRs-IB)

### 2022-04-20 manuscript/main.tex
updated manuscript

\title{An item response theory analysis of the Matrix Reasoning Item Bank (MaRs-IB)}
\shorttitle{IRT analysis of MaRs-IB}
\author{Samuel Zorowitz$^1$, Nathaniel D. Daw$^{1,2}$}
\affiliation{$^1$Princeton Neuroscience Institute, Princeton University, USA\\$^2$Department of Psychology, Princeton University, USA}

\abstract{Your abstract here.}

@@

Matrix reasoning tasks are among the most widely used measures of cognitive ability in the behavioral sciences. Much of their popularity undoubtedly reflects their versatility in use. Matrix reasoning tasks are strong (albeit impure) indicators of general intelligence \citep{gignac2015raven} and working memory capacity \citep{kane2004generality, unsworth2005working}. They are predictive of important real-world outcomes such as childhood academic performance \citep{roth2015intelligence} and performance on college entrance exams \citep{frey2004scholastic, koenig2008act}. In low-stakes testing settings, matrix reasoning tasks can additionally function as measures of motivation, willingness to expend mental effort, and other aspects of personality \citep{duckworth2011role, gignac2019maximum}. Performance on matrix reasoning tasks can also serve as an instrumental variable in studies of psychiatric populations; that is, it can be used to control for general disruptions to cognitive ability where specific domains of cognition are primarily of interest \citep{gillan2016characterizing, rouault2018psychiatric, moutoussis2021decision}.

@@

>>>>

\title{An item response theory analysis of the Matrix Reasoning Item Bank (MaRs-IB)}
\shorttitle{IRT analysis of MaRs-IB}
\author{Samuel Zorowitz$^1$, Gabriele Chierchia(?), Sarah-Jayne Blakemore(?), Nathaniel D. Daw$^{1,2}$}
\affiliation{$^1$Princeton Neuroscience Institute, Princeton University, USA\\$^2$Department of Psychology, Princeton University, USA}

\abstract{Matrix reasoning tasks are among the most widely used measures of cognitive ability in the behavioral sciences, but the paucity of matrix reasoning tests in the public domain complicates their use. Here we present an extensive investigation and psychometric validation of the matrix reasoning item bank (MaRs-IB), an open-access set of matrix reasoning items. In a first study, we calibrate the psychometric functioning of the items in the MaRs-IB in a large sample adults participants (N=1501). Using additive multilevel item structure models, we establish that the MaRs-IB has many desirable psychometric properties: its items span a wide range of difficulty, possess medium- to-large levels of discrimination, and exhibit as intended certain associations between item attributes and functioning (e.g. item complexity predicts item difficulty). However, we also find that item clones are not psychometrically equivalent and therefore not exchangeable. In a second study, we demonstrate how researchers can use the item parameters we estimated to design new MaRs-IB tests using optimal item assembly. With these methods, we design three parallel short-form measures that exhibited good score reliability and convergent validity in a second sample of adults (N=300). We hope that the materials and results made available here will encourage researchers to use the MaRs-IB in their studies.}

@@

Matrix reasoning tasks are among the most widely used measures of cognitive ability in the behavioral sciences. Much of their popularity undoubtedly reflects their versatility in use. Matrix reasoning tasks are strong (albeit impure) indicators of general intelligence \citep{gignac2015raven} and working memory capacity \citep{kane2004generality, unsworth2005working}. They are predictive of important real-world outcomes such as childhood academic performance \citep{roth2015intelligence} and performance on college entrance exams \citep{frey2004scholastic, koenig2008act}. In low-stakes testing settings, matrix reasoning tasks can additionally function as measures of motivation, willingness to expend mental effort, and other aspects of personality \citep{duckworth2011role, gignac2019maximum}. Performance on matrix reasoning tasks can also serve as an instrumental variable in studies of psychiatric populations; that is, it can be used to account for general disruptions to cognitive ability where specific domains of cognition are primarily of interest \citep{gillan2016characterizing, rouault2018psychiatric, moutoussis2021decision}.

@@

\section{Acknowledgements}
\noindent This work was funded in part by the National Center for Advancing Translational Sciences (NCATS), a component of the National Institute of Health (NIH), under award number UL1TR003017, and by an NSF Graduate Research Fellowship.
